A Classical Density-Functional Theory for Describing Water Interfaces

### 2012-05-05 papers/water-SAFT/paper.tex
various saft paper edits Tested everything on bingley.

\newcommand{\fixme}[1]{\red{[#1]}}

% needsworklater is used to annotate bits that need work, but that we
% can postpone for a while.
\newcommand{\needsworklater}[1]{\emph{[#1]}}
% needsworknow is intended to prioritize stuff that needs fixing.
\newcommand{\needsworknow}[1]{\textcolor{red}{[\emph{#1}]}}

@@

{\begin{itemize}
\color{blue}
\item Put finished things up here
\item[JRH] Rewrite abstract (copying from project)
\item[JRH] Cut the test-particle section
\item[JRH] Rewrite and join introductions
\item[JRH] Decide what to do with slit (cut)
\item[JRH] Combine and edit Other Thermo and Empirical Parameters sections
\color{red}
\item[JRH] Work on abstract
\item[DJR] Read and edit changes to abstract and introduction
\item[DJR] Look through theory section for stuff to cut
\item[DJR,JRH] Finalize figures in beginning
\item[JRH] Better/expanded description of ``macroscopic surface tension'' model in 2 rods section?
\item[JRH] Recalculate and clean up 2 rods figures and data - new code?
\item[DJR] Write text (or clean up existing) for hard sphere solutes
\item[EJK] Add 4-rods figures
\item[EJK] Add 4-rods text
\item[JRH,DJR] Rewrite conclusion
\end{itemize}
}

@@

A large fraction of interesting chemistry---including all of molecular
biology---takes place in aqueous solution.  However, while quantum
chemistry now enables us to calculate the ground state energies of
large molecules in vacuum, prediction of the free energy of even the
smallest molecules in the presence of a solvent poses a continuing
challenge due to the complex structure of a liquid and the
computational cost of \emph{ab initio} molecular
dynamics~\cite{car1985, grossman2004}.  The current state-of-the art
in \emph{ab initio} molecular dynamics is limited to around 64 water
molecules per unit cell, which places the minimum solute concentration
at 0.84 molar and allows contact between the second solvation shell of
neighboring solutes~\cite{izvekov2005, choe2007}.  On top of this,
\emph{ab initio} approaches using classical molecular dynamics have
\textcolor{red}{vague problems [DJR]}\cite{weber2010ab-initio-water}.  A
more efficient approach is needed in order to study nanoscale and
hydrophobic solutes.

@@

%\begin{figure}
%\includegraphics[width=6cm, angle=270]{figs/correlation}
%\caption{ Experimental atom-atom radial distribution functions for water at
%room temperature and pressure from reference~\cite{Soper2000}.\needsworknow{Either 
%mention this figure in the text or delete it.}  }
%\label{correlation}
%\end{figure}

@@

A number of exact properties are easily achieved in the density-functional
framework, such as the relationship between the pressure on a hard wall and
the contact density---which ensures a correct solvation free energy for
small hard solutes.  Much of the research on classical density-functional
theory has focused on the hard-sphere fluid~\cite{curtin1985,
rosenfeld1989, rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000}.
This has led to a large number of very sophisticated functionals, such as
the fundamental-measure theory (FMT) functional~\cite{rosenfeld1989,
rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000}.  This functional
\needsworknow{has many nice properties DJR}.

Several classical density functionals have already been developed for
water~\cite{ding1987, Yang1992, Jaqaman2004}, capturing some of the
qualitative behavior of water.  However, many of these are less
sophisticated than those describing the hard-sphere fluid, and a
quantitatively accurate functional describing water---which reproduces
both the macroscopic surface tension and its decrease as curvature
increases---has not yet been published.  For instance, Jaqaman
\emph{et al} developed a classical density-functional for water
including orientational effects, but their functional resulted in a
surface tension twice the experimental value, limiting it to purely
qualitative predictions~\cite{Jaqaman2004}.

\needsworknow{Add discussion and citations for more recent papers}

A recent functional presented by Lischner and Arias~\cite{lischner2010classical}
describes water as non-interacting molecules in a an
effective external potential.  This functional contains the exact
entropy for no-interacting molecules and, among other things, can 
predict the nonlinear dielectric response in water but must be paramaterized
for different temperatures.

\needsworknow{Add smooth transition from others' work to ours}

\section{Theory and Methods}
We use classical DFT to construct a new functional that will predict
the density of water molecules in a classical system.  The model we use
starts with a treatment of each individual water molecule as a purely
classical hard sphere and adds association terms based on Statistical 
Associating Fluid Theory (SAFT). Water is unlike simple
fluids (which are mainly defined by van der Waals forces and weak electrostatic
interations) because it requires direct treatment of association effects. The strong 
hydrogen bonds and polar
interactions in water are ideal for describing with SAFT, which was designed for
use with hydrogen-bonding fluids. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules. SAFT can account for the association energies
as well as any effects due to clusters or chains forming \fixme{not true}
in the liquid\cite{muller2001molecular}.  \needsworknow{Expand this a
  bit to accurately describe SAFT and its advantages.}

@@

The association functional we use is a modified version of that of Yu
and Wu\cite{yu2002fmt-dft-inhomogeneous-associating} \footnote{We
  should note that Fu and Wu\cite{fu2005vapor-liquid-dft} use almost
  the same functional, but their paper contains errors in the
  association term and is not useful as a reference.} 
which includes the effects of density inhomogeneities in the
contact value of the correlation function $g^{HS}_\sigma$. This, however, is
based on the SAFT-HS equation of state, rather than the SAFT-VR
equation of state\cite{gil-villegas-1997-SAFT-VR} which is used in the
optimal SAFT parametrization for water found by Clark \emph{et
  al}\cite{clark2006developing}.  The modification to use the SAFT-VR
equation of state involves adding a correction term, see
Equation~\ref{eq:gSW}.  \fixme{needs work}

@@

To determine the critical separation, we consider a macroscopic surface
tension model where the free energy in the macroscopic limit is given by 
\begin{equation}
F = \gamma A + pV
\end{equation}
The first term describes the surface energy and the second term
is the work needed to create a cavity of volume $V$. Since this pressure term
scales with volume, it can be neglected relative to the surface term provided
the length scale is small compared with $\frac{\gamma} p$. At room temperature,
the value of $\frac{\gamma} p$ is equal to about $20~\mu m$ which is
much larger than any of the solutes we study. We expect that for those
larger radius 
rods, the water on the sides of the 'stadium' configuration will bow inward 
between the rods and the densty will still reduce to vapor where the rods 
are closest to each other.

@@

This is consistent with the calculation of the surface tension in
Figure~\ref{figs:surface-tension} (72~mN/m) and the force per length
from Figure ~\ref{fig:rods-energy-vs-distance} (144~mN/m) which
compares the free energy per length for rods of several different
radii. This agreement indicates that even though the rods are nanoscale, the
water molecules will ``see'' them classically.  For large $d$, the rods 
simply act like two separate single rods in water and the energy will
reach an asymptote determined by the effective surface tension (shown in 
Figure~\ref{fig:energy-vs-diameter}). In figure
~\ref{fig:rods-energy-vs-distance}, we see that for small $d$ all the rods 
have the same behavior with a positive linear relationship
between free energy per length and separation during the initial state 
(where the rods have vapor between them). This indicates that the force 
per length on the rods is constant, and surprisingly independent of separation. 
By finding the slope of this curve, we calculate this force per length to be 144~mN/m.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/four-rods-energy-vs-distance}
\end{center}
\caption{ Free energy (also known as the potential of mean force) 
versus separation for four hydrophobic rods ranging in radius from
0.3~nm to 1.0~nm.
All were arbitrarily offset to zero at large separations. The
transition corresponds to the phase change from
vapor to liquid between the rods as pictured in the density profiles in 
Figure~\ref{fig:density-four-rods}. }
\label{fig:four-rods-energy-vs-distance}
\end{figure}

@@

We simulate four hydrophobic rods in water using the same radii as our
two rods ($r = 0.3$~nm~-~$1.0$~nm) and
separation distances between $d = 0$~nm to $d = 3$~nm for nearest neighbor rods. We placed
the rods in a 'square' arrangement and observed a similar behavior as the two rods
where vapor is present in the space between the rods until a critical separation
where it becomes filled with water and the rods behave like separate single rods
(shown in Fig. \ref{fig:density-four-rods}).  Free energy per length
is plotted against rod separation in Figure
\ref{fig:four-rods-energy-vs-distance} with a slight disagreement of
the predicted$2\gamma$ force per unit length.

The 'rounded square' formed by the area of the four rods and the space
inbetween them has a circumference estimated by
\begin{equation}
C_{rs} = 2 \pi r + 8r + 4d.
\end{equation}
We again equate this circumference with the total circumference of
only the four rods, which results in a critical separation of
\begin{equation}
d_{c4} = \frac{3 \pi - 4}{2} r. \label{criticalfourrods}
\end{equation}
We see in Figure \ref{fig:four-rods-energy-vs-distance} that the $r = 0.3$~nm
rod has the largest difference between simulation and prediction with
a 23\% difference, and better agreement for larger radii rods (between 0.5\% and 5.6\%).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 
% \subsection{Test-particle approach}
% 
% More interesting applications are possible when we move into three
% dimensions using spherical coordinates.  The first (but not simplest)
% thing to try is looking at the response of liquid water to a single
% fixed water molecule.  This is referred to as the \emph{test particle}
% method, and is as simple as constraining the density to
% include a delta function at the origin---meaning that a single
% molecule is fixed in location.  The free energy is then minimized
% under this constraint, giving a density profile and an associated free
% energy.  The density profile is just the self-correlation function
% multiplied by the bulk liquid density.  \fixme{We should be
%   distinguishing between the properties of the \emph{exact functional}
%   and those of an approximate functional...}  Thus, a test-particle
% calculation provides a stringent test of the correctness of any
% approximate functional.
% 
% In order to perform the test-particle test, we introduce a sharp
% gaussian density spike at the origin and relax the density under a
% constraint that this spike remain.  Figure~\ref{fig:test-particle}
% shows the resulting density profile, compared with the experimental
% $g_{\textrm{OO}}$.  Note that this data was \emph{not} used in the
% fitting of the empirical parameters of the functional.  The agreement
% between experiment and theory is both qualitatively and quantitatively
% poor.  The integral of the difference between this density profile
% (excluding the delta function) and the bulk liquid density should be
% 0.06~molecules \fixme{look up and show derivation... or cut this},
% but we find instead to be $-0.15$~molecules.  In addition, the associated
% free energy should be positive: there is an entropic cost in fixing
% one molecule in position, but for our functional the free energy of
% this density configuration is in fact negative, with a value of
% -3.5~eV, meaning that the homogeneous fluid is unstable.  This is a
% serious flaw that must be addressed in future work.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

A common model of hydrophobic solutes is the \emph{hard sphere solute}
\fixme{add citations?}.  The effect of this solute is merely to create
a cavity within which the water may not penetrate.  This is the
simplest possible solute, and will serve as a test case, which if
passed could lead to the ability to insert different chemicals into
aqueous solution, instead of a vapor cavity, in order to observe
biologically and chemically important aqueous solvation interactions.

>>>>

% needsworknow is intended to prioritize stuff that needs fixing.
\newcommand{\needsworknow}[1]{\textcolor{red}{[\emph{#1}]}}

@@

@@

A large fraction of interesting chemistry---including all of molecular
biology---takes place in aqueous solution.  However, while quantum
chemistry now enables us to calculate the ground state energies of
large molecules in vacuum, prediction of the free energy of even the
smallest molecules in the presence of a solvent poses a continuing
challenge due to the complex structure of a liquid and the
computational cost of \emph{ab initio} molecular
dynamics~\cite{car1985, grossman2004}.  The current state-of-the art
in \emph{ab initio} molecular dynamics is limited to around 64 water
molecules per unit cell, which places the minimum solute concentration
at 0.84 molar and allows contact between the second solvation shell of
neighboring solutes~\cite{izvekov2005, choe2007}.  On top of this,
standard \emph{ab initio} methods using classical molecular dynamics
without van der Waals corrections strongly over-structure water, to
the point that ice melts at over 400K\cite{yoo2009phase}!  There has
been a recent flurry of publications implicating van der Waals effects
as significant in reducing this
over-structuring\cite{lin2009importance, wang2011density,
  mogelhoj2011ab, jonchiere2011van}.  It has also been bound that
inclusion of nuclear quantum effects can have a similar
effect\cite{morrone2008nuclear}.  Each of these corrections imposes an
additional computational burden on an approach that is already
feasible only for very small number of water molecules. A more
efficient approach is needed in order to study nanoscale and
hydrophobic solutes.

@@

@@

A number of exact properties are easily achieved in the
density-functional framework, such as the relationship between the
pressure on a hard wall and the contact density---which ensures a
correct solvation free energy for small hard solutes.  Much of the
research on classical density-functional theory has focused on the
hard-sphere fluid~\cite{curtin1985, rosenfeld1989, rosenfeld1993,
  rosenfeld1997, tarazona1997, tarazona2000}.  This has led to a large
number of very sophisticated functionals, such as the
fundamental-measure theory (FMT) functional~\cite{rosenfeld1989,
  rosenfeld1993, rosenfeld1997, tarazona1997, tarazona2000,
  roth2002whitebear}.  The White Bear version of the FMT
functional\cite{roth2002whitebear}  has a number of useful
properties.  Like all FMT functionals, it is expressed as a function
of a few convolutions of the density that can be efficiently
computed.  In the homogeneous limit, it reduces to the
Carnahan-Starling equation of state, and in the strongly-confined
limit of a zero-dimensional cavity, it gives the exact free energy.

A number of classical density functionals have been developed for
water~\cite{ding1987, Yang1992, gloor2002saft, Jaqaman2004,
  clark2006developing, lischner2010classical}\needsworknow{ADD REST OF
  LIST HERE!}, capturing some of the qualitative behavior of water.
However, many of these are less sophisticated than those describing
the hard-sphere fluid, and a quantitatively accurate functional
describing water---which reproduces the temperature dependence of the
macroscopic surface tension, its decrease as curvature increases, and
its quantitative value---has not yet been published.  For instance,
the functional of Lischner \emph{et al}\cite{lischner2010classical}
treats the surface tension correctly, but can only be used at room
temperature.  A numer of classical density functionals have recently
been produced that are based on Statistical Associating Fluid Theory
(SAFT)\cite{gloor2002saft, clark2006developing}\needsworknow{add
  more}.  These functionals are based on a perturbative thermodynamic
expansion, and give a reasonable prediction for the temperature
dependence of the surface tension.  However, they also overestimate
the surface tension by \needsworknow{how much? significant factor, I
  think}, and are unsuited for studying systems with molecular-scale
inhomogeneities.

%% A recent functional presented by Lischner and Arias~\cite{lischner2010classical}
%% describe water as non-interacting molecules in a an
%% effective external potential.  This functional contains the exact
%% entropy for no-interacting molecules and, among other things, can 
%% predict the nonlinear dielectric response in water, but it must be paramaterized
%% for different temperatures.

\subsection{Statistical associating fluid theory}

Statistical Associating Fluid Theory (SAFT) is a set of models for
complex fluids in which hydrogren bonding plays a significant
role\cite{muller2001molecular}.  SAFT is used to accurately model the
equations of state of pure fluids and predict the equations of state
for mixtures over a wide range of temperatures and pressures.  SAFT is
based on Wertheim's first-order thermodynamic perturbation theory
(TPT1) \cite{wertheim1984fluidsI, wertheim1984fluidsII,
  wertheim1986fluidsIII, wertheim1986fluidsIV} which allows it to
account for strong associative interations between molecules.

The SAFT Helmholtz free energy is composed of five terms:
\begin{equation}
  F[n] = F_\textit{id} + F_\textit{hs} + F_\textit{disp} +
  F_\textit{assoc} + F_\textit{chain},
\end{equation}
where the first three terms encompass the \emph{monomer} contribution
to the free energy, the fourth is the \emph{association} free energy,
describing hydrogen bonds, and the final term is the chain formation
energy of for fluids that are chain polymers.  The final term is not
relevant for water.  While a number of SAFT formulations have been
published, we will focus on SAFT-VR\cite{gil-villegas-1997-SAFT-VR},
which has been used by Clark \emph{et al} to construct an optimal SAFT
model for water\cite{clark2006developing}.

SAFT has also been used to construct classical density functionals,
which have been applied to study the surface tension as a function of
temperature\cite{whotocite}.  Such functionals have qualitatively
predicted the dependence of surface tension on temperature, but have
in general provided poor quantitative agreement with experiment.
Moreover, the vast majority of SAFT-based classical density
functionals have made local-density approximations that make them
unsuitable for application to systems that are inhomogeneous at the
molecular length scale\cite{felipe2001examination, gloor2002saft,
  gloor2004accurate, clark2006developing, gloor2007prediction,
  kahl2008modified, gross2009density}.

\section{Theory and Methods}
We construct a new classical density functional for water.  We
start with a treatment of each individual water molecule as a purely

@@

The association functional we use is a modified version of Yu
and Wu.\cite{yu2002fmt-dft-inhomogeneous-associating} \footnote{We
  should note that Fu and Wu\cite{fu2005vapor-liquid-dft} use almost
  the same functional, but their paper contains errors in the
  association term and is not useful as a reference.} 
This functional includes the effects of density inhomogeneities in the
contact value of the correlation function $g^{HS}_\sigma$, but is
based on the SAFT-HS equation of state rather than the SAFT-VR
equation of state\cite{gil-villegas-1997-SAFT-VR}. SAFT-VR is used in the
optimal SAFT parametrization for water found by Clark \emph{et
  al}\cite{clark2006developing}, but to use it's
equation of state involves adding a correction term (see
Equation~\ref{eq:gSW}).

@@

To determine the critical separation, we start by consider a macroscopic surface
tension model where the free energy in the macroscopic limit is given by 
\begin{equation}
F = \gamma A + pV
\end{equation}
The first term describes the surface energy and the second term
is the work needed to create a cavity of volume $V$. Since this pressure term
scales with volume, it can be neglected relative to the surface term provided
the length scale is small compared with $\frac{\gamma} p$. At room temperature,
the value of $\frac{\gamma} p$ is equal to about $20~\mu m$ which is
much larger than any of the solutes we study. We expect that for those
larger radius 
rods, the water on the sides of the 'stadium' configuration will bow inward 
between the rods and the density will reduce to vapor near the center
point where the rods are closest to each other.

@@

This is consistent with the calculation of the surface tension in
Figure~\ref{fig:surface-tension} (72~mN/m) and the force per length
from Figure ~\ref{fig:rods-energy-vs-distance} (144~mN/m) which
compares the free energy per length for rods of several different
radii. This agreement indicates that even though the rods are nanoscale, the
water molecules will ``see'' them classically.  For large $d$, the rods 
simply act like two separate single rods in water and the energy will
reach an asymptote determined by the effective surface tension (shown in 
Figure~\ref{fig:energy-vs-diameter}). In Figure
~\ref{fig:rods-energy-vs-distance}, we see that for small $d$ all the rods 
have the same behavior with a positive linear relationship
between free energy per length and separation during the initial state 
(where the rods have vapor between them). This indicates that the force 
per length on the rods is constant, and surprisingly independent of separation. 
By finding the slope of this curve, we calculate this force per length to be 144~mN/m.

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/four-rods-energy-vs-distance}
\end{center}
\caption{ Free energy of hydradon 
versus separation for four hydrophobic rods ranging in radius from
0.3~nm to 1.0~nm.
All were arbitrarily offset to zero at large separations. The
transition corresponds to the phase change from
vapor to liquid between the rods as pictured in the density profiles in 
Figure~\ref{fig:density-four-rods}. }
\label{fig:four-rods-energy-vs-distance}
\end{figure}

@@

We simulate four hydrophobic rods in water using the same radii as our
two rods ($r = 0.3$~nm~-~$1.0$~nm) and
separation distances between $d = 0$~nm to $d = 3$~nm for nearest neighbor rods. We placed
the rods in a 'square' arrangement and observed a similar behavior as the two rods
where vapor is present in the space between the rods until a critical separation
where it becomes filled with water and the rods behave like separate single rods
(shown in Fig. \ref{fig:density-four-rods}).  Free energy per length
is plotted against rod separation in Figure
\ref{fig:four-rods-energy-vs-distance} with a slight disagreement of
the predicted force per unit length.

The 'rounded square' formed by the area of the four rods and the space
between them has a circumference estimated by
\begin{equation}
C_{rs} = 2 \pi r + 8r + 4d.
\end{equation}
We again equate this circumference with the total circumference of
the four separate rods, which results in a critical separation of
\begin{equation}
d_{c4} = \frac{3 \pi - 4}{2} r. \label{criticalfourrods}
\end{equation}
We see in Figure \ref{fig:four-rods-energy-vs-distance} that the $r = 0.3$~nm
rod has the largest difference between simulation and prediction with
a 23\% difference, and better agreement for larger radii rods (between 0.5\% and 5.6\%).

@@

A common model of hydrophobic solutes is the \emph{hard sphere solute}
\cite{sedlmeier2011entropy}.  The effect of this solute is merely to create
a cavity within which the water may not penetrate.  This is the
simplest possible solute, and will serve as a test case, which if
passed could lead to the ability to insert different chemicals into
aqueous solution, instead of a vapor cavity, in order to observe
biologically and chemically important aqueous solvation interactions.
